Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-03-20 paper/manuscript.Rmd
r nevents
minor edits

Forage limitation and over-grazing is a major concern for caribou populations:
it can reduce female reproductive success [@Schaefer2016] and thus could lead to
movement between islands. Forage scarcity has been proposed as a potential
reason for caribou moving between arctic islands on the sea-ice in winter
[@Miller1977]. Caribou were introduced to Fogo Island (n = 26 animals between
1964–67) and Change Island (n = 5 animals in 1964) from Newfoundland as part of
a series of translocations and introductions throughout the province (Bergerud &
Mercer 1989). It is unknown when caribou began colonizing nearby islands, but
given that some are small, over-grazing by newly arrived caribou may rapidly
deplete forage on an annual cycle [@Bergerud1990]. Although unconfirmed, it is
also possible that Fogo Island has experienced forage depletion since the
introduction of caribou in the 1960s. More likely, however, it is fine-scale
competition for foraging resources that drives caribou to swim to new islands.

Movement between islands requires caribou to assess the trade-off associated
with swimming. This trade-off can be understood as density-dependent habitat
selection governed by the Ideal Free Distribution [@Morris1987; @Bradbury2015].
Ideal Free Distribution theory predicts a fitness equilibrium: when the density
in a given habitat patch has exceeded the optimum for fitness within that patch,
animals should relocate and settle new habitat patches, so that fitness is equal
across all patches [@Bradbury2015]. Islands act as discrete habitat patches in
this case, but the costs associated with swimming create a trade-off when moving
between patches. For caribou in the Fogo Island archipelago, the role of forage
limitation in animal habitat selection patterns remains unknown, but swimming
between islands may be at least partially governed by density-dependent habitat selection.

In total, 12 of 29 collared female caribou swam between islands in the Fogo
Island archipelago. Only `r n10` of these 12 individuals had more than 10 swimming
events, suggesting that for ~10% of collared individuals swimming is relatively
common, on average occurring approximately every `r indMean` days during
ice-free seasons. We did not detect any swimming events for the remaining 17
individuals. It remains unclear whether these individuals do not swim, or
whether swimming is rare for most caribou. Overall, there appears to be three
groups of individuals: those that swim often (n = 3, every ~`r indMean` days), those that
swim occasionally (n = 9, every ~`r ind2Mean` days), and those that never swam (n = 17). We are also
unable to draw conclusions about the prevalence of swimming in the ocean by male
caribou: our initial visual observation was of an adult male, but our GPS data
only included females.

>>>>

Forage limitation and over-grazing is a major concern for caribou populations:
it can reduce female reproductive success [@Schaefer2016] and thus could lead to
movement between islands. Forage scarcity has been proposed as a potential
reason for caribou moving between arctic islands on the sea-ice in winter
[@Miller1977]. Caribou were introduced to Fogo Island (n = 26 animals between
1964–67) and Change Island (n = 5 animals in 1964) from Newfoundland as part of
a series of translocations and introductions throughout the province (Bergerud &
Mercer 1989). It is unknown when caribou began colonizing nearby islands, but
given that some are small, over-grazing by newly arrived caribou may rapidly
deplete forage on an annual cycle [@Bergerud1990]. Although unconfirmed, it is
also possible that Fogo Island has experienced forage depletion since the
introduction of caribou in the 1960s. More likely, however, is that fine-scale
competition for foraging resources drives caribou to swim to new islands.

Movement between islands requires caribou to assess the trade-off associated
with swimming. This trade-off can be understood as density-dependent habitat
selection governed by the Ideal Free Distribution [@Morris1987; @Bradbury2015].
Ideal Free Distribution theory predicts a fitness equilibrium: when the density
in a given habitat patch has exceeded the optimum for fitness within that patch,
animals should relocate and settle new habitat patches, so that fitness is equal
across all patches [@Bradbury2015]. Islands act as discrete habitat patches in
this case, but the costs associated with swimming create a trade-off when moving
between patches. For caribou in the Fogo Island archipelago, the role of forage
limitation in habitat selection patterns remains unknown, but swimming
between islands may be at least partially governed by density-dependent habitat selection.

In total, 12 of 29 collared female caribou swam between islands in the Fogo
Island archipelago. Only `r n10` of these 12 individuals had more than 10 swimming
events, suggesting that for ~10% of collared individuals swim relatively
frequently, on average occurring approximately every `r indMean` days during
ice-free seasons. We did not detect any swimming events for the remaining 17
individuals. It remains unclear whether these individuals do not swim, or
whether swimming is rare, and thus undetected, for the majority of individuals Overall, there appears to be three
groups of individuals: those that swim often (n = 3, every ~`r indMean` days), those that
swim occasionally (n = 9, every ~`r ind2Mean` days), and those that never swam (n = 17). We are also
unable to draw conclusions about the prevalence of swimming in the ocean by male
caribou: our initial visual observation was of an adult male, but our GPS data
only included females.
